---
title: "Loading Inputs: Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Inputs: Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Before running your study pipeline you need two types of inputs:

- **Cohorts** "” study populations, comparators, and outcomes as CIRCE JSON
  definitions (from ATLAS or Capr) or custom SQL
- **Concept Sets** "” phenotype definitions for diseases, exposures, covariates,
  etc.

Picard tracks both through *manifests* "” SQLite databases that record every
definition's file path, MD5 hash, metadata, and provenance. Each session you
load the manifest into memory; the SQLite file is the durable source of truth.

> For a deep-dive into the manifest architecture, derived cohorts, mid-cycle
> changes, and reset options, see the
> [Manifest: Architecture, Workflows, and Helpers](manifest_overview.html)
> vignette.

---

## The `loadingInputs.R` Template

Every picard study is generated with a `loadingInputs.R` script in
`extras/`. This script is the canonical place to run first-time setup and
subsequent-session loading. Open it in RStudio and work through the sections in
order.

The template has three top-level sections:

| Section | Purpose |
|---|---|
| **A. Concept Sets "” First Time** | `initConceptSetManifest()`, fill CSV, connect ATLAS, `importAtlasConceptSets()` |
| **B. Cohorts "” First Time** | `initCohortManifest()`, fill CSV, connect ATLAS, `importAtlasCohorts()` |
| **C. Subsequent Sessions** | `loadConceptSetManifest()` + `loadCohortManifest()` |

Run only the section that applies to your current situation.

---

## Concept Sets "” First Time

### Step 1: Initialise the manifest

```{r}
conceptSetManifest <- initConceptSetManifest()
```

This creates `inputs/conceptSets/conceptSetManifest.sqlite` if it does not
already exist and returns a `ConceptSetManifest` R6 object.

### Step 2: Create and fill the load file

```{r}
createBlankConceptSetsLoadFile()
```

Opens `inputs/conceptSets/conceptSetsLoad.csv`. Fill in one row per concept set:

| Column | Required | Notes |
|---|---|---|
| `atlasId` | Yes | ATLAS concept set ID |
| `label` | Yes | Display name |
| `domain` | Yes | OMOP domain (e.g., `drug_exposure`) |
| `sourceCode` | No | `TRUE`/`FALSE` "” whether it uses source codes |

Any additional columns are stored as tags.

### Step 3: Set up ATLAS Credentials

Before connecting to ATLAS, store your credentials securely in `~/.picard/secrets.yml` using the secrets management system:

```{r}
# Interactive setup for Atlas credentials - guides you through keyring storage
setupAtlasSecretsKeyring()

# Or edit the secrets file directly
editSecrets()
```

This creates/updates `~/.picard/secrets.yml` with your credentials stored securely. The file should contain an `atlas` section:

```yaml
atlas:
  baseUrl: "https://organization-atlas.com/WebAPI"
  authMethod: "ad"
  user: "atlas.user@company.com"
  password: !expr keyring::key_get(service = "picard", username = "atlasPassword")
```

You can store credentials three ways:
- **Plaintext** (not recommended): `password: "YourPassword"`
- **Keyring** (recommended): `password: !expr keyring::key_get(...)`
- **Environment variable**: `password: !expr Sys.getenv("ATLAS_PASSWORD")`

### Step 4: Connect to ATLAS and import

```{r}
# Credentials are automatically read from ~/.picard/secrets.yml
atlasConnection <- getAtlasConnection()

conceptSetManifest$setAtlasConnection(atlasConnection)

conceptSetManifest$importAtlasConceptSets(
  conceptSetsLoadPath = here::here("inputs/conceptSets/conceptSetsLoad.csv")
)
```

This downloads JSON definitions to `inputs/conceptSets/json/` and updates your manifest with metadata.

### Step 5: Load and review

```{r}
conceptSetManifest <- loadConceptSetManifest()
conceptSetManifest$tabulateManifest()
```

**Auto-discovery:** `loadConceptSetManifest()` scans `inputs/conceptSets/json/`
and auto-registers any `.json` files not yet in the database. Drop new concept
set JSON files there and re-run `loadConceptSetManifest()` to pick them up
without any additional import step.

---

## Cohorts "” First Time

### Step 1: Initialise the manifest

```{r}
cohortManifest <- initCohortManifest()
```

### Step 2: Create and fill the load file

```{r}
createBlankCohortsLoadFile()
```

Opens `inputs/cohorts/cohortsLoad.csv`. Fill in one row per cohort:

| Column | Required | Notes |
|---|---|---|
| `atlasId` | Yes | ATLAS cohort definition ID |
| `label` | Yes | Display name |
| `category` | Yes | Broad grouping (e.g., `"Disease Populations"`) |
| `subCategory` | No | Optional sub-grouping |

Any additional columns are stored as tags.

### Step 3: Set up ATLAS Credentials

Before connecting to ATLAS, store your credentials securely in `~/.picard/secrets.yml` using the secrets management system:

```{r}
# Interactive setup for Atlas credentials - guides you through keyring storage
setupAtlasSecretsKeyring()

# Or edit the secrets file directly
editSecrets()
```

This creates/updates `~/.picard/secrets.yml` with your credentials stored securely. The file should contain an `atlas` section:

```yaml
atlas:
  baseUrl: "https://organization-atlas.com/WebAPI"
  authMethod: "ad"
  user: "atlas.user@company.com"
  password: !expr keyring::key_get(service = "picard", username = "atlasPassword")
```

You can store credentials three ways:
- **Plaintext** (not recommended): `password: "YourPassword"`
- **Keyring** (recommended): `password: !expr keyring::key_get(...)`
- **Environment variable**: `password: !expr Sys.getenv("ATLAS_PASSWORD")`

### Step 4: Connect to ATLAS and import

```{r}
# Credentials are automatically read from ~/.picard/secrets.yml
atlasConnection <- getAtlasConnection()

cohortManifest$setAtlasConnection(atlasConnection)

cohortManifest$importAtlasCohorts(
  cohortsLoadPath = here::here("inputs/cohorts/cohortsLoad.csv")
)
```

Downloads CIRCE JSON definitions to `inputs/cohorts/json/` and records each
cohort in SQLite.

### Step 5: Load and review

```{r}
cohortManifest <- loadCohortManifest()
cohortManifest$tabulateManifest()
```

---

## Subsequent Sessions

After the first-time import, subsequent sessions only need the two load calls:

```{r}
conceptSetManifest <- loadConceptSetManifest()
cohortManifest     <- loadCohortManifest()
```

Both functions read from SQLite and rebuild the in-memory R6 objects. No
network connection or CSV file is required.

---

## What's Next

| Task | Where to go |
|---|---|
| Derived cohorts (union, subset, complement, composite) | [Manifest: Architecture, Workflows, and Helpers](manifest_overview.html) |
| Custom SQL cohorts | [Manifest: Architecture, Workflows, and Helpers](manifest_overview.html) |
| Mid-cycle changes, sync, delete, reset | [Manifest: Architecture, Workflows, and Helpers](manifest_overview.html) |
| Running the analysis pipeline | [Running the Pipeline](running_the_pipeline.html) |


