---
title: "Loading Inputs: Cohorts and Concept Sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Inputs: Cohorts and Concept Sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **Note:** This vignette is currently in development and subject to change.

## Introduction

Before running your study pipeline, you need to define the populations and phenotypes your analysis will use. Picard organizes these through two key input types:

- **Cohorts:** Define study populations, comparators, and outcomes as CIRCE-based JSON definitions or custom SQL
- **Concept Sets:** Define phenotypes for diseases, exposures, covariates, etc. as CIRCE-based JSON definitions

Picard uses *manifests* to catalog, version, and track these definitions throughout your study. This vignette walks through the complete workflow for loading and managing cohorts and concept sets.

## Manifest Overview

A manifest is a SQLite database that catalogs and tracks definitions. For each cohort or concept set, the manifest stores:

- **Metadata:** ID, label, category, source (ATLAS or manual)
- **File information:** Path and MD5 hash for change detection
- **Provenance:** When added, last modified, execution status
- **Tags:** Categorization for querying and grouping

Manifests enable reproducibility and change tracking as your study evolves.

## Working with Cohorts

The cohort workflow has four main steps:

### Step 1: Create or Update Cohorts Load File (Optional)

The `cohortsLoad.csv` file provides metadata for organizing and enriching your cohorts. You can create a blank template or edit it interactively:

**Option A: Create a blank template**

```{r eval = FALSE}
createBlankCohortsLoadFile(cohortsFolderPath = here::here("inputs/cohorts"))
```

This creates `inputs/cohorts/cohortsLoad.csv` with columns: `atlasId`, `label`, `category`, `subCategory`, `file_name`

Edit this file in Excel to add your cohort entries.

### Step 2: Import Cohort Definitions from ATLAS

#### Setting Up ATLAS Credentials

Before connecting to ATLAS, configure your credentials in your `.Renviron` file. First, view the required format:

```{r eval = FALSE}
templateatlaScredentials()
```

This shows the four credentials you need to set:

- **`atlasBaseUrl`**: Base URL to your ATLAS WebAPI (e.g., `https://organization-atlas.com/WebAPI`)
- **`atlasAuthMethod`**: Authentication method (e.g., `ad` for Active Directory)
- **`atlasUser`**: Your ATLAS username or email
- **`atlasPassword`**: Your ATLAS password

**Option A — Store in `.Renviron`**

```{r eval = FALSE}
usethis::edit_r_environ()
```

Add these lines with your actual values:

```
atlasBaseUrl='https://organization-atlas.com/WebAPI'
atlasAuthMethod='ad'
atlasUser='atlas.user@company.com'
atlasPassword='YourPassword'
```

> **Security note:** Never commit `.Renviron` to version control. Confirm it is listed in `.gitignore`.

**Option B — Store in `keyring` (recommended)**

The `keyring` package keeps credentials encrypted on your system:

```{r eval = FALSE}
# Store each credential (a prompt asks for the value)
keyring::key_set(service = "picard", username = "atlasBaseUrl")
keyring::key_set(service = "picard", username = "atlasAuthMethod")
keyring::key_set(service = "picard", username = "atlasUser")
keyring::key_set(service = "picard", username = "atlasPassword")

# Verify
keyring::key_list(service = "picard")
```

Then connect with:

```{r eval = FALSE}
atlasConn <- setAtlasConnection(useKeyring = TRUE)
```

#### Connecting and Importing

# Import specific cohorts from ATLAS
importAtlasCohorts(
  cohortsFolderPath = here::here("inputs/cohorts"),
  atlasConnection = atlasConn
)
```

This downloads JSON definitions to `inputs/cohorts/json/` and updates your manifest with metadata.

### Step 3: Load the Cohort Manifest

Load the manifest to access all cohort definitions in your study:

```{r eval = FALSE}
cm <- loadCohortManifest()
```

### Step 4: Review and Validate Manifests

Examine the cohorts that were loaded:

```{r eval = FALSE}
# View manifest list (metadata for all cohorts)
manifest_df <- cm$getManifest()

# Or get a formatted summary
summary_df <- cm$tabulateManifest()

# Query specific cohorts
cohort_1 <- cm$queryCohortsByIds(ids = 1L)
cohorts_by_tag <- cm$queryCohortsByTag(tagStrings = "category: Primary")
```

### Step 5: Apply Functions for Cohort Operations

Once manifests are loaded, you need `ExecutionSettings` to interact with the database:

```{r eval = FALSE}
# Create execution settings for a specific database
settings <- createExecutionSettingsFromConfig(configBlock = "my_cdm")

# Set execution settings on the manifest
cm$setExecutionSettings(settings)

# Now you can use database functions

# Create cohort tables in the database
cm$createCohortTables()

# Generate cohort populations
# NOTE: This step is NOT required for typical use. Cohort generation happens
# automatically during pipeline execution (via execStudyPipeline).
# This code is for demonstrating the capability only.
cm$generateCohorts()

# Get cohort counts (also typically done by pipeline)
counts <- cm$retrieveCohortCounts()
print(counts)
```

## Building Dependent Cohorts

Beyond base cohorts (imported from ATLAS or written as SQL), you can define dependent cohorts that derive from other cohorts using demographic or temporal constraints. This is useful for subpopulation analyses (e.g., males only, adults over 65, patients with prior conditions).

Dependent cohorts are built programmatically and stored in the manifest just like base cohorts.

### Temporal Subset Cohorts

Build cohorts based on temporal relationships between events. Example: "Chronic Kidney Disease (CKD) in patients with prior Type 2 Diabetes":

```{r eval = FALSE}
# Assume cm already has base cohorts loaded
# Cohort 1 = CKD, Cohort 2 = Type 2 Diabetes

ckd_given_t2d <- buildSubsetCohortTemporal(
  label = "CKD given prior T2D",
  baseCohortId = 1,          # CKD is the outcome
  filterCohortId = 2,        # Must have T2D
  temporalOperator = "before",
  temporalStartOffset = 365, # Within 1 year before CKD
  manifest = cm
)
```

**Parameters:**
- `baseCohortId`: The main population cohort
- `filterCohortId`: The condition that must occur
- `temporalOperator`: `"before"`, `"after"`, or `"overlap"` relative to base cohort
- `temporalStartOffset`: Days before/after the operator point
- `temporalEndOffset`: Optional, defines the end window

### Demographic Subset Cohorts

Build cohorts based on demographic characteristics. Example: "CKD in males only":

```{r eval = FALSE}
ckd_males <- buildSubsetCohortDemographic(
  label = "CKD Males",
  baseCohortId = 1,                  # CKD population
  genderConceptIds = c(8507),        # 8507 = Male
  manifest = cm
)
```

**Parameters:**
- `baseCohor tId`: The main population cohort
- `genderConceptIds`: Standard concept IDs for gender (e.g., 8507 for Male, 8532 for Female)
- `minAge`: Minimum age (optional)
- `maxAge`: Maximum age (optional)

Example: "CKD in adults (18+) only":

```{r eval = FALSE}
ckd_adults <- buildSubsetCohortDemographic(
  label = "CKD Adults",
  baseCohortId = 1,
  minAge = 18,
  manifest = cm
)
```

### Union and Complement Cohorts

Combine multiple cohorts or create exclusions:

**Union:** Patients in either CKD OR Diabetes:

```{r eval = FALSE}
ckd_or_t2d <- buildUnionCohort(
  label = "CKD or Type 2 Diabetes",
  cohortIds = c(1, 2),    # Union of cohorts 1 and 2
  manifest = cm
)
```

**Complement:** All patients NOT in a cohort:

```{r eval = FALSE}
no_ckd <- buildComplementCohort(
  label = "No Chronic Kidney Disease",
  populationCohortId = 1,
  excludeCohortIds = c(3),
  manifest = cm
)
```

### Stratified Cohorts

Use `buildStratifiedCohorts()` when you want to partition a cohort into **mutually exclusive subgroups** in a single call — one cohort per stratum plus an automatic **Unclassified** cohort for subjects that match none of the defined strata. This is the right tool when you need parallel subgroups for stratified analyses.

This is different from `buildSubsetCohortDemographic()`, which produces a single filtered subgroup (e.g., males aged 65+) by AND-ing all conditions together.

**Example: Split by gender**

```{r eval = FALSE}
# Produces: "CKD - Male", "CKD - Female", "CKD - Unclassified"
buildStratifiedCohorts(
  baseCohortId = 1,
  labelPrefix = "CKD",
  strata = list(
    "Male"   = list(genderConceptIds = 8507),
    "Female" = list(genderConceptIds = 8532)
  ),
  manifest = cm
)
```

**Example: Split by age group**

```{r eval = FALSE}
# Produces: "CKD - <65", "CKD - 65+", "CKD - Unclassified"
buildStratifiedCohorts(
  baseCohortId = 1,
  labelPrefix = "CKD",
  strata = list(
    "<65" = list(maxAge = 64),
    "65+" = list(minAge = 65)
  ),
  manifest = cm
)
```

**Example: Custom SQL strata (e.g., US census region)**

For partitions that cannot be expressed through standard demographic fields, supply a SQL WHERE clause string. The expression may reference `bc` (cohort table alias) and `p` (CDM person table alias):

```{r eval = FALSE}
# Produces: "CKD - West", "CKD - Midwest", "CKD - Northeast",
#           "CKD - South", "CKD - Unclassified"
buildStratifiedCohorts(
  baseCohortId = 1,
  labelPrefix = "CKD",
  strata = list(
    "West"      = "p.location_id IN (1, 4, 5, 6, 12, 14, 27, 33, 34, 38, 47, 50, 53)",
    "Midwest"   = "p.location_id IN (13, 15, 16, 17, 23, 24, 25, 29, 30, 36, 43, 51)",
    "Northeast" = "p.location_id IN (7, 20, 22, 31, 32, 35, 39, 41, 49)",
    "South"     = "p.location_id IN (2, 3, 8, 9, 10, 11, 18, 19, 21, 26, 28, 37, 42, 44, 45, 48, 52)"
  ),
  manifest = cm
)
```

**Example: Mixed demographic and custom strata**

Demographic (named list) and custom SQL strata can be combined in the same call:

```{r eval = FALSE}
buildStratifiedCohorts(
  baseCohortId = 1,
  labelPrefix = "CKD",
  strata = list(
    "Male"   = list(genderConceptIds = 8507),
    "Female" = list(genderConceptIds = 8532),
    "Rural"  = "p.location_id IN (10, 11, 23, 35)"
  ),
  manifest = cm
)
```

The Unclassified cohort is always appended automatically. Its WHERE condition is the logical negation of every named stratum, so every subject in the base cohort appears in exactly one output cohort.

### Visualizing Dependencies

Once you've defined dependent cohorts, visualize the relationship graph:

```{r eval = FALSE}
# Generate a dependency report (Mermaid diagram + table)
report <- visualizeCohortDependencies(cm)

# Optionally save to file
visualizeCohortDependencies(cm, outputPath = here::here("inputs/cohorts"))
# Creates: inputs/cohorts/cohort_dependencies.md
```

The report shows:
- Dependency structure (which cohorts depend on which)
- Mermaid diagram for visualization
- Table of all cohorts with their types

### Complete Dependent Cohort Workflow

Here's a full example combining base and dependent cohorts:

```{r eval = FALSE}
# 1. Load base cohorts from ATLAS
cm <- loadCohortManifest()

# 2. Create dependent cohorts
# Note: build* functions require the manifest and automatically register
# the new cohort — no separate addDependentCohort() call needed.
ckd_given_t2d <- buildSubsetCohortTemporal(
  label = "CKD given prior T2D",
  baseCohortId = 1,
  filterCohortId = 2,
  temporalOperator = "before",
  temporalStartOffset = 365,
  manifest = cm
)

ckd_males <- buildSubsetCohortDemographic(
  label = "CKD Males",
  baseCohortId = 1,
  genderConceptIds = c(8507),
  manifest = cm
)

ckd_adults <- buildSubsetCohortDemographic(
  label = "CKD Adults",
  baseCohortId = 1,
  minAge = 18,
  manifest = cm
)

# 3. Visualize dependencies
report <- visualizeCohortDependencies(cm)

# 4. Generate all cohorts (base + dependent)
settings <- createExecutionSettingsFromConfig(configBlock = "optum_dod")
cm$setExecutionSettings(settings)
cm$createCohortTables()
results <- cm$generateCohorts()
counts <- cm$retrieveCohortCounts()
print(counts)
```

**Note:** Dependent cohorts are generated using R code, not SQL files. When you call `cm$generateCohorts()`, Picard executes the logic for all dependent cohorts automatically. The `manifest` parameter is required by all `build*` functions — it validates parent cohorts and registers the new cohort in the manifest automatically.

## Working with Custom SQL Cohorts

Sometimes a cohort cannot be expressed as a CIRCE JSON definition from ATLAS, and is too specific or complex for a derived cohort. In these cases you can write the SQL yourself and register it as a **custom** cohort.

Custom cohorts:

- Live in `inputs/cohorts/sql/` alongside any other hand-written SQL files
- Are tracked with SQL-hash-based change detection (same as ATLAS circe cohorts)
- Are executed by `generateCohorts()` using the same standard SqlRender parameters

### SQL Requirements

Your SQL must use SqlRender parameters rather than hardcoded values. `generateCohorts()` injects these automatically:

| Parameter | Description |
|---|---|
| `@target_cohort_id` | The cohort definition ID assigned by the manifest |
| `@target_database_schema` | The schema where the cohort table resides |
| `@target_cohort_table` | The cohort table name |
| `@cdm_database_schema` | The CDM schema for referencing OMOP clinical tables |

A DELETE step before the INSERT is strongly recommended for idempotency (safe to re-run):

```sql
DELETE FROM @target_database_schema.@target_cohort_table
  WHERE cohort_definition_id = @target_cohort_id;

INSERT INTO @target_database_schema.@target_cohort_table (
  cohort_definition_id,
  subject_id,
  cohort_start_date,
  cohort_end_date
)
SELECT
  @target_cohort_id,
  person_id,
  condition_start_date,
  condition_end_date
FROM @cdm_database_schema.condition_occurrence
WHERE condition_concept_id IN (...)
;
```

### Registering a Custom Cohort

1. Place your `.sql` file in `inputs/cohorts/sql/`
2. Call `loadCohortManifest()` — the file is auto-discovered with a temporary label (the filename) and `cohortType = "circe"`
3. Call `defineCustomCohort()` to give it a proper label, tags, and mark it as `"custom"`

```{r eval = FALSE}
cm <- loadCohortManifest()

# Enrich by file path
defineCustomCohort(
  manifest = cm,
  sqlFilePath = "inputs/cohorts/sql/high_dose_steroids.sql",
  label = "High-dose corticosteroid initiators",
  tags = list(category = "Exposure", subCategory = "Corticosteroids")
)

# Or enrich by cohort ID if you already know it
defineCustomCohort(
  manifest = cm,
  cohortId = 5L,
  label = "High-dose corticosteroid initiators",
  tags = list(category = "Exposure")
)
```

The label, tags, and `cohortType = "custom"` are persisted to the SQLite manifest so they are restored automatically by future `loadCohortManifest()` calls — you only need to call `defineCustomCohort()` once per cohort.

### SQL Validation

`defineCustomCohort()` automatically checks the SQL for common portability issues and emits **warnings** (never errors) for:

- Missing required SqlRender parameters (`@target_cohort_id`, `@target_database_schema`, `@target_cohort_table`)
- No DELETE step (idempotency risk)
- INSERT missing one or more of the four required cohort columns
- Apparent hardcoded schema references

The same checks run again at execution time when `generateCohorts()` processes the cohort, so you always get a reminder even if you skipped `defineCustomCohort()`.

## Managing the Manifest Mid-Cycle

Study development is rarely linear. Cohorts get revised in ATLAS, new definitions get added mid-analysis, or old definitions are retired. Picard provides three methods to keep the manifest in sync with reality without reloading from scratch.

### Checking Manifest Health

Before syncing, get a quick status overview:

```{r eval = FALSE}
# Full status table: id, label, status, deleted_at, file_exists
cm$validateManifest()

# Summary counts
cm$getManifestStatus()
#> $active_count
#> [1] 12
#> $missing_count    # active records whose file is gone
#> [1] 1
#> $deleted_count    # soft-deleted records
#> [1] 2
#> $next_available_id
#> [1] 15
```

### Syncing Files on Disk with the Manifest

`syncManifest()` reconciles the SQLite manifest against the `json/` and `sql/` directories on disk in a single call:

- **New files** found on disk are added as new manifest entries
- **Active records** whose file has disappeared are soft-deleted (status = `'deleted'`)
- **Existing files** whose SQL hash has changed are updated in the manifest

```{r eval = FALSE}
# Sync the manifest against inputs/cohorts/json/ and inputs/cohorts/sql/
sync_results <- cm$syncManifest()

# Returns a data frame: id, label, action
# action is one of: "added", "hash_updated", "missing_flagged", "unchanged"
print(sync_results)
```

This is the go-to method when you've:

- Added new cohorts to ATLAS and re-run `importAtlasCohorts()`
- Edited a cohort SQL file directly
- Deleted a cohort file and want the manifest to reflect that

`syncManifest()` only touches `json/` and `sql/` source cohorts — derived cohorts managed by `build*` functions are left untouched.

### Removing Cohorts from the Manifest

When a cohort is no longer needed, soft-delete it to preserve the audit trail:

```{r eval = FALSE}
# Soft-delete: marks status = 'deleted', keeps the record
cm$deleteCohort(id = 5, reason = "Replaced by updated phenotype")
```

For cases where you need to permanently remove a record (e.g., a test cohort that was never used), use `permanentlyDeleteCohort()`. This requires an explicit confirmation to prevent accidents:

```{r eval = FALSE}
# Permanent removal — requires confirm = TRUE
cm$permanentlyDeleteCohort(id = 5, confirm = TRUE)
```

### Purging Deleted Cohort Data from the DBMS

Soft-deleting a cohort only updates the SQLite manifest — the cohort rows still exist in your DBMS cohort table. Use `cleanCohortTable()` to purge them:

```{r eval = FALSE}
# Requires executionSettings to be set
cm$setExecutionSettings(settings)

# Deletes rows from the DBMS cohort and checksum tables for all
# status = 'deleted' cohorts, then marks them as 'purged' in SQLite
purge_results <- cm$cleanCohortTable()
print(purge_results)   # id, label for each purged cohort
```

After `cleanCohortTable()`, those manifest records are marked `'purged'` and will not be processed again on subsequent calls — the method is safe to call repeatedly.

### Updating Cohort Labels and Tags

At any point in the study cycle you can rename a cohort or update its tags without affecting its type, SQL, or generation status. This is useful when:

- An ATLAS-imported cohort arrives with a machine-generated label you want to clean up
- You need to apply or revise categorization tags used for downstream queries or reporting
- A cohort's purpose changes and its name should reflect that

```{r eval = FALSE}
# Rename a cohort (label only — tags unchanged)
updateCohortMetadata(
  manifest = cm,
  cohortId = 3L,
  label = "Type 2 Diabetes — incident users"
)

# Add or replace tags (label unchanged)
updateCohortMetadata(
  manifest = cm,
  cohortId = 3L,
  tags = list(category = "Outcome", subCategory = "Metabolic")
)

# Update both at once
updateCohortMetadata(
  manifest = cm,
  cohortId = 3L,
  label = "Type 2 Diabetes — incident users",
  tags = list(category = "Outcome", subCategory = "Metabolic")
)
```

Changes are applied immediately to the in-memory manifest and persisted to SQLite, so they survive across sessions. Only the supplied fields are written — omitted arguments are left untouched in the database.

> **Tip:** Use `cm$tabulateManifest()` before and after to confirm the changes.

### Typical Mid-Cycle Workflow

```{r eval = FALSE}
# 1. After updating cohorts in ATLAS and re-importing
synced <- cm$syncManifest()

# 2. Review what changed
synced[synced$action != "unchanged", ]

# 3. Retire a cohort that is no longer needed
cm$deleteCohort(id = 7, reason = "Out of scope for v2")

# 4. Purge the retired cohort's data from the DBMS
cm$setExecutionSettings(settings)
cm$cleanCohortTable()

# 5. Re-generate to pick up ATLAS revisions
cm$generateCohorts()
```

## Working with Concept Sets

The concept set workflow is similar to cohorts.

### Step 1: Create or Update Concept Sets Load File (Optional)

Create metadata for organizing your concept sets:

**Option A: Create a blank template**

```{r eval = FALSE}
createBlankConceptSetsLoadFile(conceptSetsFolderPath = here::here("inputs/conceptSets"))
```

This creates `inputs/conceptSets/conceptSetsLoad.csv` with columns: `atlasId`, `label`, `category`, `subCategory`, `domain`, `sourceCode`, `file_name`

Edit this file in Excel to add your concept set entries.

### Step 2: Import Concept Set Definitions from ATLAS

Download concept set JSON definitions:

```{r eval = FALSE}
# Reuse or create ATLAS connection
atlasConn <- setAtlasConnection()

# Import concept sets from ATLAS
importAtlasConceptSets(
  conceptSetsFolderPath = here::here("inputs/conceptSets"),
  atlasConnection = atlasConn
)
```

This downloads JSON definitions to `inputs/conceptSets/json/` and updates metadata.

### Step 3: Load the Concept Set Manifest

```{r eval = FALSE}
# Load without execution settings (for review only)
csm <- loadConceptSetManifest()
```

### Step 4: Review and Validate Manifests

Examine concept sets in your study:

```{r eval = FALSE}
# Tabular view of all concept sets
manifest_df <- csm$tabulateManifest()

# Query by criteria (returns data frames)
concept_set <- csm$queryConceptSetsByIds(1L)
drug_expo <- csm$queryConceptSetsByTag(tagStrings = "domain: drug_exposure")
```

### Step 5: Apply Functions for Concept Set Operations

To extract source codes or access vocabularies, you need `ExecutionSettings`:

**Extract source codes** retrieves codes in specific vocabularies (e.g., NDC, RxNorm) for drug concepts:

```{r eval = FALSE}
# Create execution settings
settings <- createExecutionSettingsFromConfig(configBlock = "optum_dod")

# Set execution settings on the manifest
csm$setExecutionSettings(settings)

# Extract source codes for a concept set
# (e.g., NDC codes for drug concepts)
source_codes <- csm$extractSourceCodes(
  sourceVocabs = c("NDC", "RxNorm")
)
```

**Extract included codes** builds a complete list of all standard concepts included in each concept set definition and exports to Excel:

```{r eval = FALSE}
# Extract all included standard concepts from all concept sets
# Generates IncludedCodes.xlsx with one sheet per concept set
# Each sheet contains: concept_id, concept_name, vocabulary_id
csm$extractIncludedCodes(
  outputFolder = here::here("inputs/conceptSets")
)
```

## Complete Workflow Example

Here's a complete testing script combining cohorts and concept sets:

```{r eval = FALSE}
library(picard)

# === COHORTS WORKFLOW ===

# 1. Create ATLAS connection (credentials from .Renviron)
atlasConn <- setAtlasConnection()

# 2. Create load file  
createBlankCohortsLoadFile(here::here("inputs/cohorts"))

# 3. Import from ATLAS
importAtlasCohorts(
  cohortsFolderPath = here::here("inputs/cohorts"),
  atlasConnection = atlasConn
)

# 4. Load manifest and review
cm <- loadCohortManifest()
cm$getManifest()

# 5. Set up database and generate cohorts
settings <- createExecutionSettingsFromConfig(configBlock = "optum_dod")
cm$setExecutionSettings(settings)
cm$createCohortTables()
cm$generateCohorts()
counts <- cm$retrieveCohortCounts()
print(counts)

# === CONCEPT SETS WORKFLOW ===

# 1. Create load file
# Option A: Create blank template
createBlankConceptSetsLoadFile(here::here("inputs/conceptSets"))

# 2. Import from ATLAS
importAtlasConceptSets(
  conceptSetsFolderPath = here::here("inputs/conceptSets"),
  atlasConnection = atlasConn
)

# 3. Load manifest and review
csm <- loadConceptSetManifest(executionSettings = settings)
csm$getManifest()

# 4. Extract source codes (if needed)
# csm$extractSourceCodes(sourceVocabs = c("NDC"))
```

## Key Points

**Manifests as audit trail:** The manifest (SQLite database + load CSV) tracks exactly which definitions were used in each version of the study. Hash values detect if definitions change.

**Load files for metadata:** The `cohortsLoad.csv` and `conceptSetsLoad.csv` files organize and enrich definitions with human-readable metadata.

**ExecutionSettings enable database operations:** Without ExecutionSettings, you can load and review manifests. With ExecutionSettings, you can generate cohorts and extract source codes.

## Next Steps

Now that you have cohorts and concept sets loaded:

1. **Review the definitions:** Ensure your manifests contain all needed populations and phenotypes
2. **Run the pipeline:** See [Running the Pipeline](running_the_pipeline.html) to execute your first analysis
3. **Track changes:** Each manifest update via import or editor gets tracked with timestamps and hashes

## See Also

- [The Ulysses Standard Repository Structure](picard_repository_structure.html) - Where inputs fit in the project
- [Launching a Study](launching_a_study.html) - Initial setup
- [Running the Pipeline](running_the_pipeline.html) - Using cohorts in tasks
