---
title: "Setup Picard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup Picard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This guide focuses on credential management for Picard and ATLAS connections:

- Keyring-backed secrets setup
- Editing secrets with `editSecrets()`
- Reviewing and maintaining credential configuration

For builder script creation and input loading workflows, use
[Loading Inputs: Getting Started](loading_inputs.html).

---

## Configure and Edit ATLAS Credentials

Before connecting to ATLAS, configure credentials in `~/.picard/secrets.yml`.

```{r}
# Interactive setup that stores passwords in keyring
setupAtlasSecretsKeyring()

# Directly edit secrets file entries
editSecrets()
```

This creates or updates `~/.picard/secrets.yml`:

```yaml
atlas:
  baseUrl: "https://organization-atlas.com/WebAPI"
  authMethod: "ad"
  user: "atlas.user@company.com"
  password: !expr keyring::key_get(service = "picard", username = "atlasPassword")
```

Recommended: use Keyring for passwords instead of plaintext values.

---

## Credential Management Helpers

Use these helpers to inspect and maintain credential state over time.

```{r}
# Optional template helper for reference values
templateAtlasCredentials()

# Review what keyring-backed credentials are available
reviewKeyringCredentials()
```

When credentials rotate, rerun `setupAtlasSecretsKeyring()` or update values via
`editSecrets()`.

---

## Next Step

Continue with [Loading Inputs: Getting Started](loading_inputs.html) for builder
script setup, import workflows, and cohort/concept set build examples.
